Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2015-04-13 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Finish TODOs in WGBS bioinformatics analysis.

For example, in Figure \ref{fig:E18VA_M-bias} it is clear that there are problems at the start and end of both _read-1_ and _read-2_, as well as a big problem at read-position 101 in _read-2_. What is less clear is where to draw the line on the gradual decay toward the end of _read-1_ and towards the start of _read-2_. This motivated me to write a few simple functions to perform more systematic processing of M-bias results. These are included in the `MethylationTuples` R package, described in Section \ref{sec:MethylationTuples}, with the `MBias` class, its associated methods, `trim` and `plot`, and the helper function `readMBias` (__TODO: Add this class and methods to `MethylationTuples`__).

@@

The `trim` method identifies read-positions where the $rpml$ differs by more than a given value (`cutoff`) from $nrpml$. While it computes these statistics separately for each methylation type and read type, a common `cutoff` is used for all methylation types and read types. I tend to use a value of `cutoff = 0.03`, meaning that if the $median(rpml_{CpG}) = 0.75$, then any read-position with $rpml_{CpG} < 0.72$ or $rpml_{CpG} > 0.78$ is flagged as showing evidence of M-bias. It is currently left to the user to decide whether a read-position should be excluded if it displays evidence of M-bias in a single methylation type or only if it displays evidence across all methylation types. I tend to exclude read-positions that display evidence of M-bias in the methylation type I am working with, which is typically CpG methylation.

The `plot` method has a `type` argument, so that the 'raw' M-bias or the 'normalised' M-bias can be visualised (__TODO__).

```{r}
# TODO: Insert code snippet showing plots returned by `plot(MBias)` with both `type = raw` and `type = normalised`
```

>>>>

For example, in Figure \ref{fig:E18VA_M-bias} it is clear that there are problems at the start and end of both _read-1_ and _read-2_, as well as a big problem at read-position 101 in _read-2_. What is less clear is where to draw the line on the gradual decay toward the end of _read-1_ and towards the start of _read-2_. This motivated me to write a few simple functions to perform more systematic processing of M-bias results. These are included in the `MethylationTuples` R package (described in Section \ref{sec:MethylationTuples}) with the `MBias` class, its associated methods, `plot()` and `filter()`, and the helper function `readMBias()`.

@@

The `filter()` method identifies read-positions where the $rpml$ differs by more than a given value (`threshold`) from $nrpml$. While it computes these statistics separately for each methylation type and read type, a common `threshold` is used for all methylation types and read types. I tend to use a value of `threshold = 3`, meaning that if the $median(rpml_{CpG}) = 75$, then any read-position with $rpml_{CpG} < 72$ or $rpml_{CpG} > 78$ is flagged as showing evidence of M-bias[^mbias_percentage]. It is currently left to the user to decide whether a read-position should be excluded if it displays evidence of M-bias in a single methylation type or only if it displays evidence across all methylation types. I tend to exclude read-positions that display evidence of M-bias in the methylation type I am working with, which is typically CpG methylation.

[^mbias_percentage]: The M-bias files created by `bismark_methylation_extractor --mbias_only` report the methylation levels as percentages rather than proportions.
